Compositional Certified Resource Bounds\ifshort{\vspace{-.15cm}}{}

### 2014-11-03 paper/main.tex
improvements

Available techniques split in two categories.  In the
first one, tools derive impressive bounds for
imperative programs, but are not fully compositional.
Notable recent work like
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10},
and LOOPUS~\cite{SinnZV14} falls in this category.
They lack compositionality in at least three ways.
First, they all base their analysis on some form of
\emph{ranking-function}.
Since this
is often a local analysis, the loop bounds are expressed
as arithmetic expressions in terms of variables as they
are just before the loop.  This makes it hard to give a
resource bound on sequenced loops or function calls in
terms of the input parameters of a procedure.
XXX MENTION GULWANI AND KOAT XXX
Second, while all imperative programming languages
provide a function or procedure abstraction, available
tools are not able to abstract their resource behavior
and have to inline the procedure body to perform
their analysis.  Third, when such a tool fails on
program, the only way out to get a resource bound is to
rewrite the program.  There is no opportunity for sound
interaction with the tool, for example, manual proofs
cannot be composed with automatically inferred bounds.
One of the toughest challenges in the way of interaction
is that available tools work on a highly abstract version
of the source program. This abstracted program has a lot
more behaviors than the original one and is usually not
able to express memory related operations.

In the second category, tools also derive very tight
bounds on functional programs~\cite{Jost03,HoffmannAH12}.
These tools are highly compositional, scale for large
programs, and work directly on the syntax.  Despite all
these theoretical and practical advantages, it has been
a long-time open problem how to make them work in the
presence of negative integers, mutation and non-linear
control flow.  For example, the first challenge of
imperative programs is the absence of useful type system
to hook the analysis on.  Another issue is that these
tools associate a \emph{potential} directly to program
variables, doing this in imperative programs would
fail for loops as simple as \lstinline{for(i=x;i<y;i++)}
where $y - i$ decreases, but not $|i|$.
Mutation is yet another challenge, while functional
programs relate multiple fixed inputs (the typing
context) with one output (the expression typed),
imperative programs can change the value of variables
in scope, so the potential of these variables has to be
updated after mutations or unsound bounds can be derived.

This paper presents a new framework for automatically
deriving resource bounds on C programs.  This new
approach is an attempt to unify the two previous
categories: it solves the compositionality issues of
imperative world by adapting amortized-analysis based
techniques from the functional world.

Our automated analysis is able to infer linear
resource bounds on C programs with integer loops.
To our knowledge this is the first tool based on
amortized analysis able to deal with negative numbers
and differences of variables.  It is also the first
resource analysis tool for C that deals naturally with
recursive functions, resources that can be freed (e.g.
memory), and sequenced loops.  Compared to more classical
apporaches based on ranking functions, our tool inherits
the benefits of amortized reasoning and can handle using
only one simple mechanism:

\begin{itemize}
\item interactions between \emph{sequential loops or
  function calls} through size changes of variables,
\itemskipIn
\item \emph{nested loops} that influence each other
  with the same set of modified variables
\itemskipIn
\item and \emph{amortized bounds} as found, for example, in
  the Knuth-Morris-Pratt algorithm for string search.
\end{itemize}

>>>>

Available techniques split in two categories.  Tools
from the first one derive impressive bounds for
imperative programs, but they are not fully compositional.
Notable recent work like
SPEED~\cite{GulwaniMC09}, KoAT~\cite{BrockschmidtEFFG14},
PUBS~\cite{AlbertAGPZ12}, Rank~\cite{AliasDFG10},
and LOOPUS~\cite{SinnZV14} falls in this category.
They lack compositionality in at least three ways.
First, they all base their analysis on some form of
\emph{ranking-function}.
Since this
is often a local analysis, the loop bounds are
arithmetic expressions that depend on the value of
variables  just before the loop.
This makes it hard to give a
resource bound on sequenced loops or function calls in
terms of the input parameters of a procedure.
Second, while all imperative programming languages
provide a function or procedure abstraction, available
tools are not able to abstract their resource behavior
and have to inline the procedure body to perform their
analysis.  Third, when a tool fails to find a resource
bound for a program, the program has to be rewritten.
There is no opportunity for sound interaction with the
tool, for example, manual proofs cannot be composed
with automatically inferred bounds.  One reason for this
impossibility to interact is that available tools work
on a highly abstract version of the source program. This
abstracted program has a lot more behaviors than the
original one and is usually expressed in a language that
is not able to express memory operations.

In the second category, tools derive very tight
bounds on functional programs~\cite{Jost03,HoffmannAH12}.
They are highly compositional, scale for large
programs, and work directly on the syntax.  Despite all
these theoretical and practical advantages, it has been
a long-time open problem how to make them work in the
presence of negative integers, mutation and non-linear
control flow.  For example, the first challenge of
imperative programs is the absence of useful type system
to hook the analysis on.  Another issue is that these
tools associate a \emph{potential} directly to program
variables, doing this in imperative programs would
fail for loops as simple as \lstinline{for(i=x;i<y;i++)}
where $y - i$ decreases, but not $|i|$.
Mutation is yet another challenge, while functional
programs relate multiple fixed inputs (the typing
context) with one output (the expression typed),
imperative programs can change the value of variables
in scope, so the potential of these variables has to be
updated after mutations or, unsound bounds can be derived.

This paper presents a new framework for automatically
deriving resource bounds on C programs.  This new
approach is an attempt to unify the two previous
categories: it solves the compositionality issues of the
imperative world by adapting amortized-analysis based
techniques from the functional world.

Our automated analysis is able to infer linear
resource bounds on C programs with integer loops.
To our knowledge this is the first tool based on
amortized analysis able to deal with negative numbers
and differences of variables.  It is also the first
resource analysis tool for C that deals naturally with
recursive functions, resources that can be freed (e.g.
memory), and sequenced loops.  Compared to more classical
apporaches based on ranking functions, our tool inherits
the benefits of amortized reasoning.  It handles, using
only one simple mechanism:

\begin{itemize}
\item interactions between \emph{sequential loops or
  function calls} through size changes of variables,
\itemskipIn
\item \emph{nested loops} that influence each other
  with the same set of modified variables,
\itemskipIn
\item and \emph{amortized bounds} as found, for example, in
  the Knuth-Morris-Pratt algorithm for string search.
\end{itemize}
